\thetitle

### 2021-03-02 writing/challenges.tex
pass over challenges section

The rational programmer works like an algorithm. Implementing it demands a large
number of representative debugging scenarios; a reconciliation of three
different kinds of error messages; and a proper sampling strategy for the huge
space of possibilities. A coincidental challenge is the disparity of gradually
typed languages. To eliminate this variable from the comparison, this paper uses
Racket, which is thus far the only language in which all three major soundness
mechanisms are available in a reasonably robust and directly comparable manner;
for the implementation of the transient semantics, see the parallel submission
of an experience report~\cite{ttt21}.

The {\em first challenge\/} concerns the differences among the blame assignment
of the three semantic variants.  While Natural assigns blame to {\em one\/}
component, Transient assigns blame to a sequence of components. The Erasure
semantics does not blame a component {\it per se\/}, but it comes with an
exception location and a stack trace, which implicitly suggests parties to
blame.  The different strategies greatly affect the behavior of the rational
programmer (and real programmers, too).

One way to overcome this challenge is to {\em equip the rational programmer with
modes \/} that represent the alternative interpretations of blame.  For
instance, one Transient mode may pick the oldest element of a blame set as
the next component to add types because it corresponds to the earliest point in
the program's evaluation that can discover an impedance mismatch.  Another mode
may opt to treat the sequence as a stack and add types to its newest element.
When both modes are equally successful in locating an impedance mismatch,
measuring the rational programmer's effort may answer which mode is the most
effective.

Different blame information is also due to differences in the run-time checks
that the three gradual typing systems perform. This observation suggests another
mode for the rational programmer, namely, a mode that ignores the blame
information completely and instead inspects only the information in the
exceptions of the safety checks in the underlying languages.  This exception
mode serves as the {\em baseline\/} for blame's value; if the programmer in this
mode ``beats'' the others, blame systems might be useless.

The {\em second challenge\/} is to find a representative collection of
impedance mismatches. They must represent mistakes that programmers
accidentally create and that the run-time checks of academic systems catch. In
other words, the experiment calls for a collection of type mistakes in
mix-typed programs that is representative of those ``in the wild.''
Unfortunately no such collection exists, and with good reason. The kind of
mistakes needed are typically detected by unit or integration tests; even if it
takes some time to find their sources, these mistakes do not make it into code
repositories with appropriate commit messages.

@@

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutatnt component.  A
debugging scenario results from removing the types from the mutatnt. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About two thirds of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

@@

% Overcoming these challenges yields a systematic approach: 
% \begin{itemize}

% \item interesting mutators generate interesting type-level errors and thus
% debugging scenarios (section~\ref{sec:mutate});

% \item an algorithm simulates the modes of the rational programmer on these
% scenarios (section~\ref{sec:rational}); and

% \item a careful sampling of the space yields experimental
% data about the bug finding process (section~\ref{sec:sample}).
% \end{itemize}
% Getting all three aspects right is necessary to obtain a computationally
% feasible and statistically sound experiment.

% {\bf Note} A coincidental challenge is the disparity of gradually typed
% languages. While the three approaches to blame are well understood in theory,
% they have been implemented for rather different untyped languages and with
% vastly different type systems. To eliminate the effect of these differences, it
% is necessary to implement all three approaches in a single language so that an
% application of the evaluation method can use the same test programs, the same
% type system, and hence the same debugging scenarios.

% The results in this paper are for a system that builds on Greenman and
% Felleisen's effort of comparing the performance of Natural and Transient in
% Racket~\cite{gf-icfp-2018}. The extended system uses the type system of Typed
% Racket and comes with a Transient semantics that faithfully implements the
% design and the blame assignment strategy proposed by~\cite{vss-popl-2017}. The
% Erasure semantics drops the type information and runs the programs as if they
% were plain Racket programs. The next section explains this system in detail.

>>>>

At first glance implementing the methodology of
section~\ref{sec:why-rational} seems straightforward. The rational
programmer is a simple algorithm and testing whether the rational
programmer hypothesis holds amounts to running the algorithm on a
collection of scenarios.

However, three major challenges remain:  the isolation of the effect of
blame on the rational programmer across three different mechanisms; the
collection of a large number of representative debugging scenarios;  and a
proper sampling strategy for the huge space of possibilities. A
coincidental challenge is the disparity of gradually typed languages. To
eliminate this variable from the comparison, this paper uses Racket, which
is thus far the only language in which all three major soundness
mechanisms are available in a reasonably robust and directly comparable
manner; for the implementation of the transient semantics, see the
parallel submission of an experience report~\cite{ttt21}.

The {\em first challenge\/} stems from the interplay between blame and the
run-time checks of a gradual type systems. When a check fails the Natural
and Transient semantics assign blame and discard  the information in the
exception from the run-time check. Nevertheless, the exception
information may be equally or even more useful to the rational programmer
than blame. If this is the case blame may not play a critical role for the
rational programmer.\footnote{Indeed, precisely because they do not
account for confounding factors, ~\citet{lksfd-popl-2020} cannot draw any
conclusions about the pragmatics of blame despite advertisements for the
opposite. Their experiment can only conclude that so-called blame-shifting
works but not that this is because of blame.}

One way to isolate blame from this confounding factor is to {\em equip the
rational programmer with modes \/} that represent the different types of
information the rational programmer takes into account when debugging a
scenario. Specifically, the rational programmer comes with a blame and
an exception mode. Only if the former succeeds debugging a program while
the latter fails, it is safe to conclude that blame is indeed necessary
for the rational programmer. Put differently, the exception mode serves as
the {\em baseline\/} for blame's value; if the programmer in this mode
``beats'' the  blame one, a blame assignment mechanism might be useless.

Besides the run-time checks, pure luck might also amplify the usefulness
of blame. Thus a proper evaluation of blame needs to factor out any random
effects. A random rational programmer mode offers the necessary baseline
using the same ingredients as for exceptions.

Modes also offer a common vocabulary for the differences among the blame
assignment of the three semantic variants.  While Natural assigns blame to
{\em one\/} component, Transient assigns blame to a sequence of
components. The Erasure semantics does not blame a component {\it per
se\/}, but it comes with an exception location and a stack trace, which
implicitly suggests parties to blame.  The different strategies greatly
affect the behavior of the rational programmer (and real programmers,
too).

With modes, different strategy map to different ways the rational
programmer  interprets blame.  For instance, one Transient mode may pick
the oldest element of a blame set as the next component to add types
because it corresponds to the earliest point in the program's evaluation
that can discover an impedance mismatch.  Another mode may opt to treat
the sequence as a stack and add types to its newest element.  When both
modes are equally successful in locating an impedance mismatch, measuring
the rational programmer's effort may answer which mode is the most
effective.

The {\em second challenge\/} is to find a representative collection of
programs with impedance mismatches. The latter must represent mistakes that programmers
accidentally create and that the run-time checks of academic systems catch. In
other words, the experiment calls for a collection of type mistakes in
mix-typed programs that is representative of those ``in the wild.''
Unfortunately no such collection exists, and with good reason. The kind of
mistakes needed are typically detected by unit or integration tests; even if it
takes some time to find their sources, these mistakes do not make it into code
repositories with appropriate commit messages.

@@

Thus, an experimental analysis of blame needs a mostly new set of mutators.
Roughly speaking, the new mutators inject type errors into fully typed programs.
Applying such a mutator to any typed component produces a mutant component.  A
debugging scenario results from removing the types from the mutant. For the
design of such mutators, the authors relied on their own extensive programming
experience though not without discovering a major pitfall: some of their
original mutators systematically produced programs that immediately revealed the
source of the impedance mismatch. About two thirds of the original mutators
survived this first scrutiny testing. All of the remaining ones yield {\em
interesting debugging scenarios\/} (see section~\ref{sub:mutate-interesting}).

@@

### 2021-03-02 writing/rational-methodology.tex
adding methodology subsection
pass over challenges section

\item A large-scale experiment that tests the hypothesis 
  on a large number of representative scenarios.

@@

\begin{itemize}
\item The rational programmer as a procedure that consumes a program and the error message 
  the evaluation of the program produces and, uses the error message to
    select the next component of the program to add types.

@@

\item The evaluation of a blame assignment mechanism based on whether the
  information it provides was critical for the outcome of the experiment   
\end{itemize}

>>>>

  \item A large-scale experiment that tests the hypothesis on
    representative programs.

@@

  \item The rational programmer as a procedure that consumes a program and
    the error message from evaluation of the program and uses the error
    message to select the next component to add types.

@@

  \item The evaluation of a blame assignment mechanism based on whether
    the information it provides is critical for the outcome of the
    experiment.
